---
title: "Conditional Statements"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`" 
output: 
  pdf_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Creating functions and object orientated scripts are the preferred way to use R.  R functions expand the capabilities of R. By nature, R scripts are a way to organize and save data, complicated expressions, or sequences of operations for re-use.  Well configured  R functions rely on proper use of R language concepts and object orientated structures.

## R Scripts vs. R Functions
Scripts and functions have several distinguishing characteristics:

Scripts store intermediate results as persistent objects on the hard drive (e.g. in the .Data  folder), cluttering up the working database.   Functions provide more control over which results will be saved and where, be it in RAM or using swap space on the hard drive;
A function’s local or intermediate results are stored in separate evaluation frames, and pose no danger of accidentally overwriting other objects with the same names in the working database;
The dynamic memory usage of functions is usually faster than reading and writing disk files, implying functions run faster than equivalent scripts;
Functions more easily support modular and vectorized code development;
Unlike scripts, functions can be deployed as a group by packaging them as an R library.
Types of R Functions
There are three broad types of R functions objects:

Standard Call Functions: these functions are user defined or intrinsic to base R.  Examples include mean(), lm(), and summary();
Infix Operators: These are standard operators for data manipulation, transformation and customization.  Examples include +, –, %*%, etc.
Replacement Functions: These non-arithmetic functions act on data objects and extend manipulation to include physical redefinition.  Examples include diag(), length(), etc.
Meanwhile, each of these function types can be defined as employing S3 or S4 methods.  In simple terms, the difference is a matter of development standards, and impacts which R kernel or interpreter is used.

## The Function Body
The syntax and generalized body of any user-defined function is defined below:

function.name <- function(argument.list) {
 expressions
}

A function can have no arguments, one or many. Multiple argument inputs must be separated by commas.  When providing a default value for an argument, the syntax argument=value must be used in the argument list.

The body of the function is made of up of one or more R expressions, which are legal commands. The use of function bracketing { } is optional in the case of a short, one-line expressions, but is necessary when more than one command is specified. Functions can be defined within the body of other functions.  Finally, it is good practice to include comments in the function body.

The value returned by a function is the last value of the function body, which is usually an unassigned expression.  Alternatively, a function may be terminated at any stage by calling the return() function.  If return() is given several named arguments, then the value returned is a list object with several (named) components.

## Example of Simple Function Writing in R
A function is presented to illustrate R language conventions and formatting styles:

```{r}
# Define the function
summy2 <- function(x,y){
  2*(x+y)
}
summy2(3,5)


summ <− function(x) {
   x.mean <− mean(x)
   x.var <− var(x)
   x.sum <− sum(x)
   out <− list(mean = x.mean,var = x.var,sum = x.sum)
   return(out)
}
x <- c(1, 2, 3, 4, 5)
summ(x)
y <- unlist(summ(x))
```
## if else if

```{r ifels}
a <- 3
if (a > 2){
  print(" a is larger than 2")
}

x <- 5
if (x>5)
{
  y <- x
}else{
  y <- 5*2
}
y

a <- 3
if (a==2){
  print("a is 2")
}else if (a==3){
  print ("a is 3")
}else if (a ==4){
  print("a is 4")
}else{
  print("a is something else")
}
```
## swith 
The switch() command is conceptually similar to an if-clause. Ho wever, the difference is that one can test more than one condition at the same time. For instance, in the example below, the switch() command tests 3 conditions, because it has 3 executable components, indicated by the “{ }” environments. If the variable “a” is 1, the first commands are executed, if “a” is 2 the second, and so on.
```{r swith}
a <- 3
switch (a,{print("A");print("B")},
        {print("C")},
        {print("D")}
)

```

##LOOPS
In R, there are two different ways to realize a looping behavior. The first is by using a for-loop, and the second by using a while-loop. A looping behavior means the consecutive execution of the same procedure for a number of steps. The number of steps can be fixed, or variable.

```{r}
Fib <- rep(1,100)
for (i in 3:100){
  Fib [i] <- Fib[i-1]+Fib[i-2]
}

Fib

i <- 0
while (i<3){
  i <- i+1 
  print(i)
}
```

## Break
Both loop functions can be interrupted at any time during the execution of the loop using the break() command. Frequently, this is used in combination with an if-clause within a loop to test for a specific decision that shall lead to the interruption of the
loop.

```{r break}
for (i in 1:3){
  print(i)
  if (i==2) break
}

```

## Repeat-Loop
For completeness, we want to mention that there is actually a third type of loop in R,the repeat-loop. However, in contrast with a for-loop and a while-loop, this does not come with an interruption condition, but is in fact an infinite loop that does never stop. For this reason, the repeat() command needs to be used always in combination with the break() statement:

```{r Repeat-loop}
i <- 0
repeat{
  i <- i+1
  print(i)
  if (i > 10) break
}

```
## Exercise
y <- c(rep(100,6),80,82,85,95,100,100,60,70,75,85,98,100,20,30,50,70,80,100,10,20,40,60,75,100,0,10,40,60,70,100)
data <-  matrix(y,nrow = 6,byrow = TRUE)
data

