---
title: "Data_cleaning"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`" 
output: 
  pdf_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Cleaning data
The processes of cleaning your data can be the most time-consuming part of any data analysis. The most important steps are considered below. While there are many approaches, those using the dplyr and tidyr packages are some of the quickest and easiest to learn.

Package	Function	Use
dplyr	select	select variables/columns
dplyr	filter	select observations/rows
dplyr	mutate	transform or recode variables
dplyr	summarize	summarize data
dplyr	group_by	identify subgroups for further processing
tidyr	gather	co nvert wide format dataset to long format
tidyr	spread	convert long format dataset to wide format
Examples in this section will use the starwars dataset from the dplyr package. The dataset provides descriptions of 87 characters from the Starwars universe on 13 variables.

#Selecting variables
The select function allows you to limit your dataset to specified variables (columns).
```{r}
library(dplyr)

# keep the variables name, height, and gender
newdata <- select(starwars, name, height, gender)

# keep the variables name and all variables 
# between mass and species inclusive
newdata <- select(starwars, name, mass:species)

# keep all variables except birth_year and gender
newdata <- select(starwars, -birth_year, -gender)

```
# Selecting observations
The filter function allows you to limit your dataset to observations (rows) meeting a specific criteria. Multiple criteria can be combined with the & (AND) and | (OR) symbols.

```{r}
library(dplyr)

# select females
newdata <- filter(starwars, sex == "female")

# select females that are from Alderaan
newdata <- filter(starwars, 
                  sex == "female" & 
                  homeworld == "Alderaan")


# select individuals that are from 
# Alderaan, Coruscant, or Endor
newdata <- filter(starwars, 
                  homeworld == "Alderaan" | 
                  homeworld == "Coruscant" | 
                  homeworld == "Endor")

# this can be written more succinctly as
newdata <- filter(starwars, 
                  homeworld %in% c("Alderaan", "Coruscant", "Endor"))

```

#Creating/Recoding variables
The mutate function allows you to create new variables or transform existing ones.
```{r}
library(dplyr)

# convert height in centimeters to inches, 
# and mass in kilograms to pounds
newdata <- mutate(starwars, 
                  height = height * 0.394,
                  mass   = mass   * 2.205)

# if height is greater than 180 
# then heightcat = "tall", 
# otherwise heightcat = "short"

newdata <- mutate(starwars,
                  heightcat = ifelse(height > 180,"tall","short"))
                  
# convert any eye color that is not 
# black, blue or brown, to other
newdata <- mutate(starwars, 
                  eye_color = ifelse(eye_color %in% c("black", "blue",  
                                                      "brown"),
                                     eye_color,"other"))
# set heights greater than 200 or 
# less than 75 to missing
newdata <- mutate(starwars, 
                  height = ifelse(height < 75 | height > 200, NA,height))
```
#Summarizing data
The summarize function can be used to reduce multiple values down to a single value (such as a mean). It is often used in conjunction with the by_group function, to calculate statistics by group. In the code below, the na.rm=TRUE option is used to drop missing values before calculating the means.

```{r}
library(dplyr)

# calculate mean height and mass
newdata <- summarize(starwars, 
                     mean_ht = mean(height, na.rm=TRUE), 
                     mean_mass = mean(mass, na.rm=TRUE))
newdata

# calculate mean height and weight by gender
newdata <- group_by(starwars, sex)
newdata <- summarize(newdata, 
                     mean_ht = mean(height, na.rm=TRUE), 
                     mean_wt = mean(mass, na.rm=TRUE))
newdata

```
# Using pipes
Packages like dplyr and tidyr allow you to write your code in a compact format using the pipe %>% operator. Here is an example.
The %>% operator passes the result on the left to the first parameter of the function on the right.
```{r}
newdata <- filter(starwars, 
                  sex == "female")
newdata <- group_by(species)
newdata <- summarize(newdata, 
                     mean_ht = mean(height, na.rm = TRUE))

# this can be written as
newdata <- starwars %>%
  filter(sex == "female") %>%
  group_by(species) %>%
  summarize(mean_ht = mean(height, na.rm = TRUE))

```
#  Reshaping data
Some graphs require the data to be in wide format, while some graphs require the data to be in long format.

```{r}
id <- c(1,2,3)
name <- c("Bill","Bob","Mary")
sex <- c("Male","Male","Female")
age <- c(22,25,18)
income <- c(5500,7500,9000)
wide_data <- data.frame(id,name,sex,age,income)

library(tidyr)
long_data <- gather(wide_data, 
                    key="variable", 
                    value="value", 
                    sex:income)
#Conversely, you can convert a long dataset to a wide dataset using
wide_data <- spread(long_data, variable, value)
```
#Missing data
Real data are likely to contain missing values. There are three basic approaches to dealing with missing data: feature selection, listwise deletion, and imputation. Let’s see how each applies to the msleep dataset from the ggplot2 package. The msleep dataset describes the sleep habits of mammals and contains missing values on several variables.
Feature selection
In feature selection, you delete variables (columns) that contain too many missing values.
```{r}
data(msleep, package="ggplot2")

# what is the proportion of missing data for each variable?
pctmiss <- colSums(is.na(msleep))/nrow(msleep)
round(pctmiss, 2)
#Sixty-one percent of the sleep_cycle values are missing. You may decide to drop it.
```
#Listwise deletion
Listwise deletion involves deleting observations (rows) that contain missing values on any of the variables of interest.

```{r}
# Create a dataset containing genus, vore, and conservation.
# Delete any rows containing missing data.
newdata <- select(msleep, genus, vore, conservation)
newdata <- na.omit(newdata)

```
# Imputation
Imputation involves replacing missing values with “reasonable” guesses about what the values would have been if they had not been missing. There are several approaches, as detailed in such packages as VIM, mice, Amelia and missForest. Here we will use the kNN function from the VIM package to replace missing values with imputed values.

```{r}
# Impute missing values using the 5 nearest neighbors
library(VIM)
newdata <- kNN(msleep, k=5)
```
Basically, for each case with a missing value, the k most similar cases not having a missing value are selected. If the missing value is numeric, the mean of those k cases is used as the imputed value. If the missing value is categorical, the most frequent value from the k cases is used. The process iterates over cases and variables until the results converge (become stable). This is a bit of an oversimplification - see Imputation with R Package VIM for the actual details